---
title: 'Intro to R: Week 5'
output: pdf_document
---
```{r Setup, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, results="hide", 
                      message=FALSE, warning=FALSE, eval=FALSE)
```

#Topics Covered: Date/Time Data and Data Manipulation

Before we begin, you will need to install several packages.  Please do so before
class to make sure that everything is working properly.

```{r Install, eval=FALSE, echo=TRUE}

install.packages("lubridate")
install.packages("reshape2")
install.packages("dplyr")
install.packages("ggplot2")

```

\begin{center}
\large{\textbf{Part 1: Dealing with Date/Time Data}}
\end{center}

**Task 1: Introducing the `lubridate` package**

*Step 1.1* Load the lubridate library.  Use `?lubridate` to see what
this package has to offer.

```{r Task1.1}

library("lubridate")

?lubridate

```

*Step 1.2* What is the current system time?  What is the current time in London?
Hint: use `now()` and `with_tz()`.

```{r Task1.2}

now()

class(now()) # this is a time class object

with_tz(now(), tz="GMT") # Use OlsonNames() to see all possible timezones

```

*Step 1.3* On what day of the week were you born?  Hint: use `ymd()` and `wday()`.
On what day of the week was your 21st birthday?  Hint: use `years()`.

```{r Task1.3}

bday <- ymd("1988 10 23", # This could also be 1988-10-23 or 1988 10 23 etc.
            tz="EST5EDT") # I was born on the East coast
                         
wday(bday, label=TRUE) # Note that this function is returning an ordered factor
                       # without label=TRUE, the returned value is numeric

bday.21 <- bday + years(21)

wday(bday.21, label=TRUE)

```

**Task 2: Importing, formatting, and binning data from the SIO Pier**

*Step 2.1* Read in the shore stations data from the SIO pier, plot the temperature 
measurements, then add a column with a POSIX timestamp. Hint: use `paste()` and 
`ymd_hms()`.

```{r Step2.1}

setwd("~/Desktop/IntroR/Week 5/")

# I downloaded these data directly from http://www.sccoos.org/query/
# SCOOS supplies header metadata which we don't want, so I set skip=6
# The header info did specify that all times are provided in UTC

sio.pier <- read.csv("Shore Stations_Jan-Apr 2015.csv", 
                     skip=6, header=TRUE, stringsAsFactors=FALSE)

plot(sio.pier$Water.Temperature..C., type="l")

# Stick all of the date time information together into a single string
tstring <- paste(sio.pier$year, sio.pier$month, sio.pier$day, sio.pier$time)

# Convert the string to a POSIX class object
timestamp <- ymd_hms(tstring, tz="UTC")

# Now add that information back into the data frame
sio.pier$Timestamp <- timestamp

```

*Step 2.2* Use `plot()` or `ggplot()` to plot time v. temperature for the SIO
pier record.

```{r Step2.2, eval=FALSE}

plot(sio.pier$Timestamp, sio.pier$Water.Temperature..C., type="l")

library(ggplot2)
ggplot(data=sio.pier, 
       aes(x=Timestamp, y=Water.Temperature..C.)) +
  geom_line()

```

*Step 2.3* These measurements were taken every six minutes, and they're noisy.
Write a function that uses `interval()` and `within()` to bin data by a specified
number of hours.  Hint: the structure of your function will be very similar to 
the binning function we created last week for CTD data.

```{r Step2.3}

binPier <- function(data, bin.length){
  
  # Function to average SIO Pier measurements into bins of specified no. of hrs
  # data is a data frame of SCOOS Pier data 
  # bin.size is the number of hours data should be binned into
  # The function returns a data frame with the timestamp and averaged measurements
  
  df <- data
  
  # convert the date/time data into a POSIX timestamp column
  df$Timestamp <- ymd_hms(paste(df$year, df$month, df$day, df$time), tz="UTC")
  
  # create a sequence of time bins from the start to the end of the data record
  t.bins <- seq(range(df$Timestamp)[1], 
                range(df$Timestamp)[2], 
                by=bin.length*60*60) # need to specify this in secs, not hrs
  
  df.out <- data.frame()
  
  for (t in 1:(length(t.bins)-1)){
    
    int <- interval(t.bins[t], t.bins[t+1]) 
    
    i <- which(df$Timestamp %within% int)
    
    # note that you could also use 
    # i <- which(df$Timestamp >= t.bins[t] & df$Timestamp < t.bins[t+1])
    
    nd <- data.frame("Bin.Start"=t.bins[t],
                     "Avg.Temp"=mean(df$Water.Temperature..C.[i]),
                     "Avg.Chl"=mean(df$Chlorophyll..ug.L.[i]),
                     "Avg.Sal"=mean(df$Salinity..PSU.[i]))
    
    df.out <- rbind(df.out, nd)
    
  } # end t
  
  return(df.out)
  
} # end binPier

# Run the function with a bin length of 6 hrs
binned.data <- binPier(sio.pier, 6)

```

*Step 2.4* Now calculate the average temperature in 24 hour bins and plot the 
day of the year vs. temperature using plot or ggplot2. Hint: use `yday()`.

```{r Step2.4, eval=FALSE}
# Run the function with a bin length of 24 hrs
pier.24 <- binPier(sio.pier, 24)

# Plotting

plot(yday(pier.24$Bin.Start), pier.24$Avg.Temp, type="l")

ggplot(pier.24, aes(x=yday(Bin.Start), y=Avg.Temp))+
  geom_line()

```

\newpage
\begin{center}
\large{\textbf{Part 2: Data Manipulation}}
\end{center}

There are whole books written about data manipulation (also called data wrangling)
in R, so this will be a very very brief overview to introduce you to a few of the 
packages and functions available.  

**Task 3: Handy data manipulation functions in base R**

*Step 3.1* Bin your SIO pier data by 1 hr, then use the `apply()` and `fivenum()`
function to return the minimum, first quartile, median, third quartile, and maximum
values of average temperature, average chlorophyll, and average salinity from your
binned SIO pier dataset.

```{r Task3.1}

pier.1hr <- binPier(sio.pier, 1) 

pier.summary <- apply(pier.1hr[,2:4], 2, fivenum)

```

This is a very minimal example of the `apply` family of functions.  Basically, 
anything that you can do with a `for` loop you can also do with `apply`.  

*Step 3.2* Use `table()` to determine how many measurements were collected in each 
month of the SIO pier record.

```{r Task3.2}

table(month(sio.pier$Timestamp, label=TRUE))

```


**Task 4: Using the `reshape2` package**

*Note* Many of the functions in the `reshape2` package have analogs in the `dplyr`
package.  I'm more familiar with the `reshape2` package so that's what I'll 
present here, but if you're interested in data wrangling definitely check
out `dplyr`!

*Step 4.1* Use the `melt()` and `dcast()` functions to generate a data frame
of average temperature, salinity, and chlorophyll by month using your binned one
hour data frame.  This should not take more than two lines of code.  *Bonus:*
Use ggplot to create a three-panel timeseries plot of temperature, chlorophyll, 
and salinity. 

```{r Task4.1, eval=FALSE}

library(reshape2)

# Melt the data usin Bin.Start as the identifying variable
m.pier <- melt(pier.1hr, id.var="Bin.Start")
# now, each row = one observation

# Now recast the data with months in the rows and measurements in columns, 
# using mean() to calculate the monthly averages.  Note that because there are
# missing values, we have to include na.rm=TRUE

mean.pier <- dcast(m.pier, 
                   month(Bin.Start, label=TRUE)~variable, 
                   mean, na.rm=TRUE)

names(mean.pier)[1] <- "Month" # Tidy up the data frame

### Plotting

ggplot(m.pier, aes(x=Bin.Start, y=value, color=variable))+
  geom_line()+
  facet_wrap(~variable, ncol=1, scales="free_y")

# Note: yes, there is something funky about the chlorophyll measurements!
  
```

*Step 4.2* Read in the islands dataset and create a data frame of average percent
hard corals, soft corals, etc. by island.  Again, manipulating the data should not
take more than two lines of code. *Bonus* Use `ggplot2` to create pie charts
of average cover on each island. 

```{r Task4.2, eval=FALSE}

# Read in the islands dataset
islands <- read.csv("IslandsData.csv", header=TRUE, stringsAsFactors=FALSE)

# Melt the data, using both the Island and the Quadrat as identifiers
l.islands <- melt(islands, id.vars=c("Island", "Quadrat"))

# Now recast the data with Island as rows and cover type in columns
avg.islands <- dcast(l.islands, Island~variable, mean)

### Plotting

# re-melt so we can plot the averages
long.avg <- melt(avg.islands)

ggplot(long.avg, aes(x="", y=value, fill=factor(variable)))+
  geom_bar(stat="identity", width=1)+
  coord_polar(theta="y")+
  facet_wrap(~Island)

```

**Task 5: Regular Expressions and Partial String Matching**

*Step 5.1* Load the file SampleIndexes.csv and extract the samples which come from 
the cruise with identifier 1311COFI.  Hint: use `grep()`.

```{r Task5.1}

# Read in the data
indexes <- read.csv("SampleIndexes.csv", header=TRUE, stringsAsFactors=FALSE)

# Find the location of the samples from the CalCOFI cruise
grep("1311COFI", indexes$LABID)

# Use that information to pull out the CalCOFI samples
COFI <- indexes[grep("1311COFI", indexes$LABID),]

```

*Step 5.2* Which of the CalCOFI sample indexes begin with the sequence ATC?

```{r Task5.2}

grep("^ATC", COFI$INDEX)

# note that if you use grep("ATC", COFI$INDEX) the answer is different
# because grep will look for the pattern anywhere in the string rather than
# just at the beginning

```

*Step 5.3* Were any samples labeled with indexes GCCGCG, GGGCCA, or CATTTT? Hint:
use `%in%` or `match()`. 
```{r Task5.3}

match(c("GCCGCG", "GGGCCA", "CATTTT"), COFI$INDEX)

c("GCCGCG", "GGGCCA", "CATTTT") %in% COFI$INDEX

```
